\bf Visual Inference for Significance and Goodness-of-Fit Testing for a Social Network Model

### 2017-10-25 ASApaper/paper-child.Rnw
couple of comments - sorry have to go blind with the Latex

\section{Introduction}
% \st{
% \begin{enumerate}
% \item significance testing: what is it and why is it important
% \item goodness-of-fit testing: what is it and why is it important
% \item difficulty of these two for sna, why do vis inf
% \end{enumerate}
% }
Two of the most imporant pieces of statistical modeling are significance testing of model parameters and goodness-of-fit measures and tests. In the former, the data are usually assumed to come from a simple model under the null hypothesis, and additional parameters are added and tests done to determine if the additional should be included in the model. In the latter, the model of interest is examined to determine how well it fits the data. Both of these aspects of statistical modeling increase greatly in difficulty as the complexity of the model of interest increases. The more complicated the model, the harder it is to determine fit or to include or exclude a parameter.

Some particularly complicated sets of models are those designed to model network change. A \textit{network} is any set of things, such as people, computers, or neurons, that are connected in some way, through social relations, internet connection, or electrical impulses in the brain. We refer to the ``things" in the network as \textit{nodes}, or \textit{actors} in a social network, and the connections as \textit{edges}, or \textit{ties} in a social network. Network objects are particularly difficult to model due to the dependencies inherent to the data. This complexity can increases when, instead of looking at one instance of a network, we consider the dynamics of network change in between those singular observed instances. This type of modelling for dynamic networks is often performed on social network data, such as for friendship networks among students or the spread of HIV in communities of men sleeping with other men. These models lack the asymptotics required to perform many well-known goodness-of-fit tests, and the maximum likelihood estimation of parameters is so difficult that it can make significance testing significant as well \citep{goldenberg09}.

We propose new methods for significance and goodness-of-fit testing for a set of social network models, stochastic actor-oriented models for dynamic network data \citep{saompaper}. These new methods use \textit{visual inference} in place of traditional statistical methods for social network models, such as Wald tests for significance of parameters and in- and outdegree distribution metrics for determining goodness-of-fit. By using the visual inference methodology of \citet{Bujaetal}, we can look at the entire dataset simulated from a network model as opposed to a single metric on the network such as outdegree, or a $p$-value for a single parameter in the model.

The paper is outlined as follows: Section 2 gives a basic overview of visual inference and the lineup protocal. Section 3 provides an introduction to the our models of interest, stochastic actor-oriented models. Sections 4 and 5 discuss problems with traditional significance tests and goodness-of-fit measures for these models and how visual inference methods can be used as solutions. Section 6 outlines our experimental protocol, and we close with a discussion in Section 7.

\section{Visual Inference}

Data visualizations are an important component of data analysis, providing a mechanism for discovering patterns in data. Pioneering research by \citet{gelman:2004}, \citet{Bujaetal} and \citet{majumder:2011} provide methods to quantify the significance of discoveries made from visualizations. \citet{Bujaetal} introduced two protocols, the Rorschach and the lineup protocol, which bridge the gulf between traditional statistical inference and exploratory data analysis. %The Rorschach protocol consists of a set of $m$ (usually, $m=20$) plots (called the {\it null plots}) rendered from data that is consistent with a given null model. The Rorschach protocol helps to understand the extent of randomness in the null model. 
We use line lineup protocol. Under this protocol, a plot of the observed data is placed randomly among a set of $m-1$ null plots, and human observers are then asked to examine the lineup and to identify the most different plot. If observers identify the data plot, this is quantifiable evidence against the null hypothesis.

@@

%To model family that we apply to the Senate data is the Stochastic Actor-Oriented Model (SAOM) family. 
This family of models for dynamic network data, which were first introduced by \citet{saompaper}, incorporates both network structure and node-level information to describe how a network observed on two or more occasions changes in time. The two titular properties of SAOMs, stochasticity and actor-orientation, are crucial to understanding networks as they exist naturally: social networks are ever-changing as relationships decay or grow in seemingly random ways, and most actors in them have characteristics that could affect how they change their ties to other nodes in the network. 
% Talk about SAOMs
These unique properties allow for the fitting of some very complicated models to inherently complex data, so it can be exceedingly difficult to interpret parameters and their corresponding estimates. The sheer amount of possible parameters to include in the model combined with the difficulty of interpretation make parameter selection and goodness-of-fit testing burdensome as well.

Broadly, a SAOM takes network structure and node covariate information into account in two ways and models the network changes as a continuous time Markov chain (CTMC). First, the rate of change between states is dictated by a rate function that describes \textit{when} changes in the network occur, and secondy, the objective function describes \textit{what} those state changes are. As in many other network models, the variables of interest, $x_{ij}$, are the binary edges of the network. For any two nodes $i \neq j$ in the network, $i,j \in \{1, 2, \dots, n=\text{the number of nodes}\}$, 
  \begin{equation}\label{eq:edgevars}
  x_{ij} =
  \begin{cases}
                                   1 & \text{if an edge from $i$ to $j$ exists} \\
                                   0 & \text{otherwise}
  \end{cases}
  \end{equation}
The edges are also treated as \textit{directed}, so it is not required that $x_{ij} = x_{ji}$, and self-referencing edges or loops, $x_{ii}$, are not allowed. The entire network at time point $m = 1, 2, \dots M$ for $M$ the number of network observations, is denoted $x(t_m)$. More detail on SAOMs follows, and additional details can be found in \citet{saompaper, snijders01, snijders2010, snijdersetall:2007, snijders:2010, snijders:2017},

\subsection{Rate Function}

All changes in SAOMs are treated as changes made by the nodes, or \textit{actors}, in the network. So, each actor, $i$, gets a chance to make a change according to the rate function, typically denoted $\lambda_i$, which dictates when relationships between nodes in the network can change. In general, the rate function can take the network structure e.g. outdegree of node $i$, and the node covariates into account, but we use the simple rate function, which is constant over all nodes in a given time period. We denote the rate from $t_{m}$ to $t_{m+1}$ as $\alpha_m$ for $m = 1, \dots, M-1$. Using this notation, the waiting time to the next chance for actor $i$ to make a change is exponentially distributed with expected value $\alpha_m^{-1}$. Since the rate is the same for all actors, the waiting time for \textit{any} actor to get the chance to change is exponentially distributed with expected value $(n\alpha_m)^{-1}$.

\subsection{Objective Function}

>>>>

\section{Introduction}
% \st{
% \begin{enumerate}
% \item significance testing: what is it and why is it important
% \item goodness-of-fit testing: what is it and why is it important
% \item difficulty of these two for sna, why do vis inf
% \end{enumerate}
% }
Two of the most important pieces of statistical modeling are significance testing of model parameters and goodness-of-fit tests. In the former, the data are usually assumed to come from a simple model under the null hypothesis, and additional parameters are tested whether they significantly contribute to explaining variability in the data. In the latter, the model of interest is examined to determine how well it fits the data. Both of these aspects of statistical modeling increase greatly in difficulty as the complexity of the model of interest increases. 
%The more complicated the model, the harder it is to determine fit or to include or exclude a parameter.

Some particularly complicated sets of models are those designed to model network change. A \textit{network} is any set of things, such as people, computers, or neurons, that are connected in some way, through social relations, internet connection, or electrical impulses in the brain. We refer to the ``things" in the network as \textit{nodes}, or \textit{actors} in a social network, and the connections as \textit{edges}, or \textit{ties} in a social network. 
Dependencies inherent to the data make network objects particularly difficult to model. Even more challenging is the situation when we go beyond single instances of a network and consider the dynamics of network change between observed instances. This type of modelling for dynamic networks is often performed on social network data, such as for friendship networks among students or the spread of HIV in  sexually active communities.% of men sleeping with other men. 
These models lack the asymptotics required to perform many well-known goodness-of-fit tests, and the maximum likelihood estimation of parameters is so difficult that it can make significance testing significant as well \citep{goldenberg09}. 
\hh{is the previous sentence a quote? I'm not sure that I follow.}

We propose new methods for significance and goodness-of-fit testing for a set of social network models, stochastic actor-oriented models for dynamic network data \citep{saompaper}. Specifically, we are using \textit{visual inference} in place of traditional statistical methods for social network models, such as Wald tests for significance of parameters and in- and outdegree distribution metrics for determining goodness-of-fit. Visual inference, introduced by \citet{Bujaetal}, allows us to look at the entire dataset simulated from a network model as opposed to a (set of) usually one-dimensional metric(s) derived from the network such as outdegree, or a $p$-value for a single parameter in the model.

The paper is outlined as follows: Section~\ref{vi} gives a basic overview of visual inference and the lineup protocol. Section 3 provides an introduction to the our models of interest, stochastic actor-oriented models. Sections 4 and 5 discuss problems with traditional significance tests and goodness-of-fit measures for these models and how visual inference methods can be used as solutions. Section 6 outlines our experimental protocol, and we close with a discussion in Section 7.

\hh{where do the results come in?}

\section{Visual Inference}\label{vi}

Data visualizations are an important component of data analysis, providing a mechanism for discovering patterns in data. Pioneering research by \citet{gelman:2004}, \citet{Bujaetal} and \citet{majumder:2011} provide methods to quantify the significance of discoveries made from visualizations. \citet{Bujaetal} introduced two protocols, the Rorschach and the lineup protocol, which bridge the gulf between traditional statistical inference and exploratory data analysis. %The Rorschach protocol consists of a set of $m$ (usually, $m=20$) plots (called the {\it null plots}) rendered from data that is consistent with a given null model. The Rorschach protocol helps to understand the extent of randomness in the null model. 
Here, we use the lineup protocol. Under this protocol, a plot of the observed data is placed randomly among a set of $m-1$ null plots (where $m =20$, usually), and human observers are then asked to examine the lineup and to identify the most different plot. If an observer identifies the data plot, this is quantifiable evidence against the null hypothesis. \hh{Similarly, an observer has a chance of 1 in $m$ to pick the data plot from the lineup by simply guessing, ie in a situation where the data plot is virtually indistinguishable from the null plots.}

@@

%To model family that we apply to the Senate data is the Stochastic Actor-Oriented Model (SAOM) family. 
Stochastic Actor-Oriented Models (SAOMs) are a family of models for dynamic network data \citep{saompaper} that incorporate both network structure and node-level information to describe how a network observed on two or more occasions changes over time. The two titular properties of SAOMs, stochasticity and actor-orientation, are crucial to understanding networks as they exist naturally: social networks are ever-changing as relationships decay or grow in seemingly random ways, and most actors in them have characteristics that could affect how they change their ties to other nodes in the network. 
% Talk about SAOMs
These unique properties allow for the fitting of some very complicated models to inherently complex data, so it can be exceedingly difficult to interpret parameters and their corresponding estimates. The sheer amount of possible parameters to include in the model combined with the difficulty of interpretation make parameter selection and goodness-of-fit testing burdensome as well.

Broadly, a SAOM takes network structure and node covariate information into account in two ways and models the network changes as a continuous time Markov chain (CTMC). First, the rate of change between states is dictated by a rate function that describes \textit{how often} changes in the network occur, and secondly, the objective function describes \textit{what} those state changes are. As in many other network models, the variables of interest,  are the binary edges of the network. Let $x_{ij}$ denote the edge between nodes $i$ and $j$, where $i,j \in \{1, 2, \dots, n=\text{the number of nodes}\}$. 
$x_{ij}$ is modelled as a binary variable, i.e.
  \begin{equation}\label{eq:edgevars}
  x_{ij} =
  \begin{cases}
                                   1 & \text{if an edge from $i$ to $j$ exists} \\
                                   0 & \text{otherwise}
  \end{cases}
  \end{equation}
Edges are treated as \textit{directed}, i.e. in general $x_{ij} \neq x_{ji}$, and self-referencing edges or loops are not allowed, i.e.  $x_{ii} = 0$ for all $i$. 
Assume, the network is observed $M$ times at time points  $t_1 < t_2 < ... < t_M$, then the  entire network at time point $t_m$ is denoted as $x(t_m)$. In sections~\ref{rf} and~\ref{of} we discuss the rate and objective functions of a SOAM in more depth.  Additional details on SOAMs can be found in \citet{saompaper, snijders01, snijders2010, snijdersetall:2007, snijders:2010, snijders:2017},

\subsection{Rate Function}\label{rf}

All changes in SAOMs are treated as changes made by the nodes, or \textit{actors}, in the network, i.e.\ each actor, $i$, gets a chance to make a change according to the rate function, typically denoted $\lambda_i$, which dictates when relationships between nodes in the network can change. In general, the rate function can take the network structure e.g. outdegree of node $i$, and the node covariates into account, but we use the simple rate function, which is constant over all nodes in a given time period. We denote the rate from $t_{m}$ to $t_{m+1}$ as $\alpha_m$ for $m = 1, \dots, M-1$. Using this notation, the waiting time to the next chance for actor $i$ to make a change is exponentially distributed with expected value $\alpha_m^{-1}$. Since the rate is the same for all actors, the waiting time for \textit{any} actor to get the chance to change is exponentially distributed with expected value $(n\alpha_m)^{-1}$.

\subsection{Objective Function}\label{of}
